\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-10-20 doc/USER_MANUAL/05_running_the_solver.tex
improved the note on the acoustic formulation

However, this formulation assumes that the density is spatially constant, or locally slowly varying (e.g. inside a given spectral element),
because the use of a potential for acoustic waves is justified by the fact that acoustic displacements are irrotational.
In the cases for which the background density has strong local variations, that formulation is not valid any more.
Indeed, taking the rotational of \eqref{potential}
\begin{eqnarray*}
\nabla \times (\rho \mathbf{u}) &=& \nabla\times\nabla \chi
\end{eqnarray*}
Since for any scalar $\chi$, $ \nabla\times\nabla \chi = 0$, this yields
\begin{eqnarray*}
\nabla\rho\times\mathbf{u} + \rho\nabla\times\mathbf{u} &=& \mathbf{0}\\
\nabla\times\mathbf{u}&=&-(1/\rho)\nabla\rho\times\mathbf{u}
\end{eqnarray*}
Then, one notices that in the case of strong density gradients the rotational of $\mathbf{u}$ is no longer zero.
Thus, the potential formulation \eqref{potential} is no more valid. It is valid only when $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible,
i.e. when both the gradient of density and the value of displacement are small, or if the displacement is parallel or quasi-parallel to the gradient of density,
in which case their rotational is also close to zero.
\newline

Another approach exists for acoustic waves, based on a displacement potential instead of on a potential of $\rho \mathbf{u}$:
\begin{equation}
\mathbf{u} = \nabla \chi \, ,
\end{equation}
which gives $\nabla\times\mathbf{u} = \mathbf{0}$.
But in this case the gradient of density is neglected in the Eulerian momentum equation in order to identify gradient terms.
Indeed, from the Eulerian momentum equation for inviscid acoustic waves only
\begin{eqnarray*}
\rho\partial_t^2\mathbf{u} &=& -\nabla p\\
\nabla (\partial_t^2\chi)&=& -(1/\rho)\nabla p \, .
\end{eqnarray*}
When $\nabla\rho \gg \mathbf{0}$ strictly speaking one cannot identify $\partial_t^2\chi$ with $p$.

>>>>

However, this formulation indirectly assumes that the density is spatially constant, or locally slowly varying (e.g. inside a given spectral element),
because acoustic displacements must be irrotational. Since we know that $\nabla \times \mathbf{u}$ must be equal to zero
we can see that when density has strong local variations, equation (\ref{potential}) is not valid any more.
Indeed, taking the rotational of \eqref{potential}
\begin{eqnarray*}
\nabla \times (\rho \mathbf{u}) &=& \nabla\times\nabla \chi \, ,
\end{eqnarray*}
since the curl of a gradient is always zero i.e. for any scalar $\chi$, $ \nabla\times\nabla \chi = 0$, we get
\begin{eqnarray*}
\nabla\rho\times\mathbf{u} + \rho\nabla\times\mathbf{u} &=& \mathbf{0}\\
\nabla\times\mathbf{u}&=&-(1/\rho)\nabla\rho\times\mathbf{u}
\end{eqnarray*}
Then, one notices that in the case of strong density gradients the rotational of $\mathbf{u}$ is no longer zero.
Thus, the potential formulation \eqref{potential} is no more valid. It is valid only when $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible,
i.e. when both the gradient of density and the value of displacement are small, or if the displacement is parallel or quasi-parallel to the gradient of density,
in which case their rotational is also close to zero.
\newline

Another approach exists for acoustic waves, based on a displacement potential instead of on a potential of $\rho \mathbf{u}$:
\begin{equation}
\mathbf{u} = \nabla \chi \, ,
\end{equation}
which correctly gives $\nabla\times\mathbf{u} = \mathbf{0}$.
But in this case the gradient of density is neglected in the Eulerian momentum equation in order to identify gradient terms.
Indeed, from the Eulerian momentum equation for inviscid acoustic waves only
\begin{eqnarray*}
\rho\partial_t^2\mathbf{u} &=& -\nabla p\\
\nabla (\partial_t^2\chi)&=& -(1/\rho)\nabla p \, .
\end{eqnarray*}
When $\nabla\rho \gg \mathbf{0}$ strictly speaking one cannot identify $\partial_t^2\chi$ with $p$.
